The Tornado Project: Counting Tornadoes
========================================================

A number of studies have analyzed the Tornado data available from the NOAA [Storm Prediction Center](http://www.spc.noaa.gov/gis/svrgis/). Below is an attempt to reproduce the statistics presented by the Storm Prediction Center and three other studies in the recent literature.

First, load the required libraries and read the data.

```{r}
source("lib_torn.R")

torn <- read_torn_data()
```

Reproduce stats from SPC. Tornadoes by year and month since 1950 and also number of fatalities and injuries. Compare below numbers with [stats from SPC](http://www.spc.noaa.gov/archive/tornadoes/ustdbmy.html). For most of the years and months, the numbers below exactly match those from SPC.

```{r}
out_spc <- rep_stats_SPC(torn)
lapply(out_spc, head)

lapply(out_spc, tail)
```

Reproduce stats from Boruff et al 2003. Compare below with Table 1 and 2 from Boruff et al. Although not an exact match, the numbers below are pretty close to those from Boruff et al.

```{r}
out_boruff <- rep_stats_Boruff(torn)
lapply(out_boruff, head)
```

Reproduce stats from Verbout et al 2006. Compare below with numbers from Figure 1 of Verbout et al 2006. Below numbers match well with those in Figure 1 of Verbout (visual comparison) and those described in the first paragraph of Page 88 of Verbout et al.

```{r}
out_verbout <- rep_stats_Verbout(torn)
head(out_verbout, 10)
tail(out_verbout, 10)
```

Reproduce stats from Simmons et al 2013. Compare below with Table 2 and Table 3 from Simmons et al 2013. Most likely, Simmons et al in Table 2 have made an error. I am not sure how the tornadoes were countd. Moreover, the "Max" values in Table 2 cannot be greater than the upper bounds on damage of each of the bins (see Table 1 of Simmons et al for damage bin intervals).

```{r}
out_simmons <- rep_stats_Simmons(torn)
options(scipen = 10)
lapply(out_simmons, head, 9)
```

