Harold Brooks and others used data from the Storm Prediction Center to estimate the probability of occurrence of a tronado day near any location in the mainland United States (Brooks, Harold E., Charles A. Doswell III, and Michael P. Kay. "Climatological estimates of local daily tornado probability for the United States." Weather and Forecasting 18.4 (2003): 626-640.). 

The below code produces numbers consistent, if not identical, to those from Brooks et al. The discrepancy could be due to the differences in Tornado data used (due to regular quality control the latest data used here may not be identical to the one used by SPC).

Read raw data add column names similar to those used by [Elsner et al 2013](http://rpubs.com/jelsner/4205) and consistent with [documentation](http://www.spc.noaa.gov/gis/svrgis/). 

```{r}
library(plyr)
library(ggplot2)

torn <- read.csv("data/1950-2012_torn.csv", 
                 header = FALSE, 
                 sep = ",", 
                 as.is = TRUE)

colnames(torn) <- c("OM", "YEAR", "MONTH", "DAY", "DATE", "TIME", "TIMEZONE", 
                    "STATE", "FIPS", "STATENUMBER", "FSCALE", "INJURIES", 
                    "FATALITIES", "LOSS", "CROPLOSS", "SLAT", "SLON", "ELAT", 
                    "ELON", "LENGTH", "WIDTH", "NS", "SN", "SG", "F1", "F2", 
                    "F3", "F4")
```

Tornadoes spanning multiple counties are listed separately for each county. Thus, a single tornado could appear multiple times. Identify unique tornadoes using YEAR, OM and NS. Check for uniqueness; especially, tornadoes spanning multiple years (i.e, those which begin on 12/31 and end on 1/1); need to check only those with NS > 1.

```{r}
dec31 <- subset(torn, MONTH == 12 & DAY == 31 & NS != 1)
jan01 <- subset(torn, MONTH == 1 & DAY == 1 & NS != 1)
if (nrow(dec31) > 0 & nrow(jan01) > 0) {
  stop("check! unique id assignment may not be accurate!")
}
torn$id <- paste(torn$YEAR, torn$MONTH, torn$OM, torn$NS, sep = "-")
```

Brooks et al used data from 1955-99 and excludd Alaska, Hawaii and Puerto Rico. 

```{r}
torn <- subset(torn, YEAR %in% seq(1955, 1999))
torn <- subset(torn, !(STATE %in% c("AK", "HI", "PR")))

str(torn)

head(torn)
tail(torn)
```

Function to summarize counts of unique number of tornadoes by year and month.

```{r}
count_unique_tornadoes <- function(in_df, monthly_stats = TRUE) {
  
  require(plyr)
  
  if (monthly_stats) {
    # some months dont have data; assign NAs to those months
    mon_totals <- expand.grid(MONTH = seq(1, 12), stringsAsFactors = FALSE)
    
    # number of unique tornadoes per month
    mon_torn <- ddply(.data = in_df, 
                      .variables = .(MONTH),
                      .fun = function(x_df) length(unique(x_df$id)), 
                      .drop = FALSE)
    
    mon_totals <- merge(mon_totals, mon_torn, by = "MONTH", all = TRUE)
    
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)), mon_totals$V1)
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique", month.abb)  
  } else {
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)))
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique")      
  }
  
  return (out_mat)
}
```

Counts of tornadoes by year, similar to those in Figure 1 of Brooks et al.

```{r}
event_stats <- ddply(.data = torn, 
                     .variables = .(YEAR), 
                     .fun = count_unique_tornadoes,
                     monthly_stats = FALSE)

str(event_stats)

event_stats
```

Graphic similar to Figure 1 of Brooks et al.

```{r}
gfx_line <- ggplot(data = event_stats, aes(x = YEAR, y = N_unique))
gfx_line <- gfx_line + geom_point()
gfx_line <- gfx_line + ylim(c(0, 1600))
gfx_line <- gfx_line + stat_smooth(method = lm, se = FALSE)
gfx_line <- gfx_line + xlab("Year") + ylab("Tornadoes")
print(gfx_line)
```

Number of torado days per year, similar to those in Figure 2 of Brooks et al.

```{r}
day_stats <- unique(torn$DATE)
day_stats <- do.call("rbind", strsplit(day_stats, "-"))
head(day_stats)

day_stats <- as.data.frame(table(day_stats[, 1]), stringsAsFactors = FALSE)
colnames(day_stats) <- c("YEAR", "Days")

head(day_stats)
```

Graphic similar to Figure 2 of Brooks et al.

```{r}
gfx_line <- ggplot(data = day_stats, aes(x = YEAR, y = Days, group = 1))
gfx_line <- gfx_line + geom_point()
gfx_line <- gfx_line + ylim(c(0, 250))
gfx_line <- gfx_line + stat_smooth(method = lm, se = FALSE)
gfx_line <- gfx_line + xlab("Year") + ylab("Tornado Days")
print(gfx_line)
```

