Harold Brooks and Charles Doswell used data from the Storm Prediction Center as well as tornado data from other countries to analyze tornadoes by damage category (Brooks, Harold, and Charles A. Doswell. "Some aspects of the international climatology of tornadoes by damage classification." Atmospheric research 56.1 (2001): 191-201.). 

The below code produces numbers consistent, if not identical, to those from Brooks and Doswell. The discrepancy could be due to the differences in Tornado data used (due to regular quality control the latest data used here may not be identical to the one used by SPC).

Read raw data add column names similar to those used by [Elsner et al 2013](http://rpubs.com/jelsner/4205) and consistent with [documentation](http://www.spc.noaa.gov/gis/svrgis/). 

```{r}
library(plyr)
library(ggplot2)
library(reshape2)

torn <- read.csv("data/1950-2012_torn.csv", 
                 header = FALSE, 
                 sep = ",", 
                 as.is = TRUE)

colnames(torn) <- c("OM", "YEAR", "MONTH", "DAY", "DATE", "TIME", "TIMEZONE", 
                    "STATE", "FIPS", "STATENUMBER", "FSCALE", "INJURIES", 
                    "FATALITIES", "LOSS", "CROPLOSS", "SLAT", "SLON", "ELAT", 
                    "ELON", "LENGTH", "WIDTH", "NS", "SN", "SG", "F1", "F2", 
                    "F3", "F4")
```

Tornadoes spanning multiple counties are listed separately for each county. Thus, a single tornado could appear multiple times. Identify unique tornadoes using YEAR, OM and NS. Check for uniqueness; especially, tornadoes spanning multiple years (i.e, those which begin on 12/31 and end on 1/1); need to check only those with NS > 1.

```{r}
dec31 <- subset(torn, MONTH == 12 & DAY == 31 & NS != 1)
jan01 <- subset(torn, MONTH == 1 & DAY == 1 & NS != 1)
if (nrow(dec31) > 0 & nrow(jan01) > 0) {
  stop("check! unique id assignment may not be accurate!")
}
torn$id <- paste(torn$YEAR, torn$MONTH, torn$OM, torn$NS, sep = "-")

str(torn)

head(torn)
tail(torn)
```

Function to summarize counts of unique number of tornadoes by year and month.

```{r}
count_unique_tornadoes <- function(in_df, monthly_stats = TRUE) {
  
  require(plyr)
  
  if (monthly_stats) {
    # some months dont have data; assign NAs to those months
    mon_totals <- expand.grid(MONTH = seq(1, 12), stringsAsFactors = FALSE)
    
    # number of unique tornadoes per month
    mon_torn <- ddply(.data = in_df, 
                      .variables = .(MONTH),
                      .fun = function(x_df) length(unique(x_df$id)), 
                      .drop = FALSE)
    
    mon_totals <- merge(mon_totals, mon_torn, by = "MONTH", all = TRUE)
    
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)), mon_totals$V1)
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique", month.abb)  
  } else {
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)))
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique")      
  }
  
  return (out_mat)
}
```

### Figure 1

Counts of annual average tornadoes by decade per FSCALE category, produced by Brooks and Doswell in Figure 1 of their paper. 

Identify decades. Ignore FSCALE of -9 (unknown). Data prior to 1950 is not available. Ignore decade of 2010 because of fewer data points.

```{r}
time_breaks <- c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2020)
time_labels <- c("1950s", "1960s", "1970s", "1980s", "1990s", "2000s", "2010s")
  
torn$time_cat <- cut(torn$YEAR, 
                     breaks = time_breaks,                         
                     labels = time_labels,                         
                     include.lowest = TRUE,                        
                     right = FALSE)

events_by_scale <- ddply(.data = subset(torn, YEAR %in% c(1950:2009) & 
                                          FSCALE %in% c(0:5)),                     
                         .variables = .(FSCALE, time_cat),                     
                         .fun = count_unique_tornadoes,                    
                         monthly_stats = FALSE)

events_by_scale
```

To get annual average per decade divide unique tornadoes in the decade by 10.

```{r}
gfx_line <- ggplot(data = events_by_scale, 
                   aes(x = FSCALE, y = N_unique/10, group = time_cat))
gfx_line <- gfx_line + geom_line(aes(colour = time_cat))
gfx_line <- gfx_line + geom_point(aes(colour = time_cat))
gfx_line <- gfx_line + ylab("Number") + scale_y_log10(breaks = 10**seq(-1, 3))
gfx_line <- gfx_line + theme(legend.title = element_blank())
gfx_line <- gfx_line + ggtitle("Average Annual Number of Tornadoes by decade, 1950-2009")
png("gfx_brooks2001_fig1.png")
plot(gfx_line)
garbage <- dev.off()
```
![figure 1] [brooks1]


### Figure 3

Counts of tornadoes by FSCALE and region. Consistent with Brooks and Doswell, split mainland US into four regions. Discard non-mainland USA regions.

```{r}
levels(as.factor(torn$STATE))
torn <- subset(torn, !(STATE %in% c("AK", "HI", "PR")))
Fn_Assign_Region <- function(state_name) {
  if (state_name %in% c("OK", "KS", "NE")) {
    return ("Central US")
  } else if (state_name %in% c("FL")) {
    return ("Florida")
  } else if (state_name %in% c("WA", "OR", "CA", "MT", "ID", "NV", "UT", "CO", 
                               "WY", "NM", "AZ", "TX", "ND", "SD")) {
    return ("Western US")
  } else {
    return ("Eastern US")
  }
}

torn$Region <- sapply(torn$STATE, FUN = Fn_Assign_Region, USE.NAMES = FALSE)
```

Tornadoes by FSCALE and Region. Set F2 counts to 100, consitent with Brooks and Doswell, and scale the counts in the other categories accordingly.

```{r}
events_by_region <- ddply(.data = subset(torn, YEAR %in% c(1950:1995) & 
                                          FSCALE %in% c(0:5)),                                              
                          .variables = .(FSCALE, Region),                                                                        
                          .fun = count_unique_tornadoes,                                                                       
                          monthly_stats = FALSE)

events_by_region

events_by_region <- dcast(events_by_region, Region ~ FSCALE, value.var = "N_unique")
region_names <- events_by_region$Region
events_by_region <- events_by_region[, c(-1)] * (100 /events_by_region[, "2"])
events_by_region$Region <- region_names
events_by_region <- melt(events_by_region, id.vars = c("Region"))

events_by_region
```

Graphic similar to Figure 3 of Brooks and Doswell.

```{r}
gfx_line <- ggplot(data = events_by_region, 
                   aes(x = variable, y = value, group = Region))
gfx_line <- gfx_line + geom_line(aes(colour = Region))
gfx_line <- gfx_line + geom_point(aes(colour = Region))
gfx_line <- gfx_line + ylab("Number") + scale_y_log10(breaks = 10**seq(-1, 3))
gfx_line <- gfx_line + xlab("FSCALE") 
gfx_line <- gfx_line + theme(legend.title = element_blank())
gfx_line <- gfx_line + ggtitle("Number of Tornadoes, per 100 F2s, by region, 1950-1995")
png("gfx_brooks2001_fig3.png")
plot(gfx_line)
garbage <- dev.off()
```
![figure 3] [brooks3]

Brooks and Doswell considered only data from 1950-1995. Here is the graphic with data from 1950-2012.

```{r}
events_by_region <- ddply(.data = subset(torn, YEAR %in% c(1950:2012) & 
                                          FSCALE %in% c(0:5)),                                              
                          .variables = .(FSCALE, Region),                                                                        
                          .fun = count_unique_tornadoes,                                                                       
                          monthly_stats = FALSE)

events_by_region <- dcast(events_by_region, Region ~ FSCALE, value.var = "N_unique")
region_names <- events_by_region$Region
events_by_region <- events_by_region[, c(-1)] * (100 /events_by_region[, "2"])
events_by_region$Region <- region_names
events_by_region <- melt(events_by_region, id.vars = c("Region"))

gfx_line <- ggplot(data = events_by_region, 
                   aes(x = variable, y = value, group = Region))
gfx_line <- gfx_line + geom_line(aes(colour = Region))
gfx_line <- gfx_line + geom_point(aes(colour = Region))
gfx_line <- gfx_line + ylab("Number") + scale_y_log10(breaks = 10**seq(-1, 3))
gfx_line <- gfx_line + xlab("FSCALE") 
gfx_line <- gfx_line + theme(legend.title = element_blank())
gfx_line <- gfx_line + ggtitle("Number of Tornadoes, per 100 F2s, by region, 1950-2012")
png("gfx_brooks2001_fig3_new.png")
plot(gfx_line)
garbage <- dev.off()
```
![figure 33] [brooks33]


[brooks1]: gfx_brooks2001_fig1.png "figure 1"
[brooks3]: gfx_brooks2001_fig3.png "figure 3"
[brooks33]: gfx_brooks2001_fig3_new.png "figure 33"
