Harold Brooks and others used data from the Storm Prediction Center to estimate the probability of occurrence of a tronado day near any location in the mainland United States (Brooks, Harold E., Charles A. Doswell III, and Michael P. Kay. "Climatological estimates of local daily tornado probability for the United States." Weather and Forecasting 18.4 (2003): 626-640.). 

The below code produces numbers consistent, if not identical, to those from Brooks et al. The discrepancy could be due to the differences in Tornado data used (due to regular quality control the latest data used here may not be identical to the one used by SPC).

Read raw data add column names similar to those used by [Elsner et al 2013](http://rpubs.com/jelsner/4205) and consistent with [documentation](http://www.spc.noaa.gov/gis/svrgis/). 

```{r}
library(plyr)
library(ggplot2)

torn <- read.csv("data/1950-2012_torn.csv", 
                 header = FALSE, 
                 sep = ",", 
                 as.is = TRUE)

colnames(torn) <- c("OM", "YEAR", "MONTH", "DAY", "DATE", "TIME", "TIMEZONE", 
                    "STATE", "FIPS", "STATENUMBER", "FSCALE", "INJURIES", 
                    "FATALITIES", "LOSS", "CROPLOSS", "SLAT", "SLON", "ELAT", 
                    "ELON", "LENGTH", "WIDTH", "NS", "SN", "SG", "F1", "F2", 
                    "F3", "F4")
```

Tornadoes spanning multiple counties are listed separately for each county. Thus, a single tornado could appear multiple times. Identify unique tornadoes using YEAR, OM and NS. Check for uniqueness; especially, tornadoes spanning multiple years (i.e, those which begin on 12/31 and end on 1/1); need to check only those with NS > 1.

```{r}
dec31 <- subset(torn, MONTH == 12 & DAY == 31 & NS != 1)
jan01 <- subset(torn, MONTH == 1 & DAY == 1 & NS != 1)
if (nrow(dec31) > 0 & nrow(jan01) > 0) {
  stop("check! unique id assignment may not be accurate!")
}
torn$id <- paste(torn$YEAR, torn$MONTH, torn$OM, torn$NS, sep = "-")
```

Brooks et al used data from 1955-99 and excludd Alaska, Hawaii and Puerto Rico. 

```{r}
torn <- subset(torn, YEAR %in% seq(1955, 1999))
torn <- subset(torn, !(STATE %in% c("AK", "HI", "PR")))

str(torn)

head(torn)
tail(torn)
```

Function to summarize counts of unique number of tornadoes by year and month.

```{r}
count_unique_tornadoes <- function(in_df, monthly_stats = TRUE) {
  
  require(plyr)
  
  if (monthly_stats) {
    # some months dont have data; assign NAs to those months
    mon_totals <- expand.grid(MONTH = seq(1, 12), stringsAsFactors = FALSE)
    
    # number of unique tornadoes per month
    mon_torn <- ddply(.data = in_df, 
                      .variables = .(MONTH),
                      .fun = function(x_df) length(unique(x_df$id)), 
                      .drop = FALSE)
    
    mon_totals <- merge(mon_totals, mon_torn, by = "MONTH", all = TRUE)
    
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)), mon_totals$V1)
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique", month.abb)  
  } else {
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)))
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique")      
  }
  
  return (out_mat)
}
```

### Figure 1
Counts of tornadoes by year, similar to those in Figure 1 of Brooks et al.

```{r}
event_stats <- ddply(.data = torn, 
                     .variables = .(YEAR), 
                     .fun = count_unique_tornadoes,
                     monthly_stats = FALSE)

str(event_stats)

event_stats
```

Graphic similar to Figure 1 of Brooks et al.

```{r}
gfx_line <- ggplot(data = event_stats, aes(x = YEAR, y = N_unique))
gfx_line <- gfx_line + geom_point()
gfx_line <- gfx_line + ylim(c(0, 1600))
gfx_line <- gfx_line + stat_smooth(method = lm, se = FALSE)
gfx_line <- gfx_line + xlab("Year") + ylab("Tornadoes")
print(gfx_line)
```

### Figure 2
Number of torado days per year, similar to those in Figure 2 of Brooks et al.

```{r}
day_stats <- unique(torn$DATE)
day_stats <- do.call("rbind", strsplit(day_stats, "-"))
head(day_stats)

day_stats <- as.data.frame(table(day_stats[, 1]), stringsAsFactors = FALSE)
colnames(day_stats) <- c("YEAR", "Days")

head(day_stats)
```

Graphic similar to Figure 2 of Brooks et al.

```{r}
gfx_line <- ggplot(data = day_stats, aes(x = YEAR, y = Days, group = 1))
gfx_line <- gfx_line + geom_point()
gfx_line <- gfx_line + ylim(c(0, 250))
gfx_line <- gfx_line + stat_smooth(method = lm, se = FALSE)
gfx_line <- gfx_line + xlab("Year") + ylab("Tornado Days")
print(gfx_line)
```


### Figure 3, Temporal Smoothing

Estimate probability of a tornado day for a leap year, consistent with Brooks et al. Estimate the number of a time a day of the year (doy) occurs in the period 1980-1999. Use 2012 as a reference leap year for the inclusion of Feb 29. One could accomplish the below task using "chron" or "lubridate" package; but, for now I am trying to avoid the importing of these packages.

```{r}
torn_8099 <- subset(torn, YEAR >= 1980 & YEAR <= 1999)
doy <- unique(torn_8099$DATE)
doy <- substr(doy, 6, 10)
doy <- paste0("2012-", doy)
ref_Dec31_2011 <- as.numeric(as.Date("2011-12-31"))
doy_freq <- as.numeric(as.Date(doy)) - ref_Dec31_2011
str(doy_freq)
```

Convert doy_freq to probability and check for probs outside of 0-1 interval. Also prob for Feb 29 should be based on number of leap years during the data time period and not the entire time span, consistent with Brooks et al.

```{r}
time_span <- 1999 - 1980 + 1
doy_prob <- as.data.frame(table(doy_freq), stringsAsFactors = FALSE)
colnames(doy_prob) <- c("doy", "freq")
# fix for Feb 29, 5 leap years in 1980-1999
doy_prob$span <- ifelse(doy_prob$doy != 60, time_span, 5)
doy_prob$prob <- doy_prob$freq / doy_prob$span

# add 0s for missing days
miss_days <- data.frame(doy = as.character(seq(1:366)),
                        date = seq(as.Date("2012-01-01"), as.Date("2012-12-31"), by = "day"),
                        stringsAsFactors = FALSE)
doy_prob <- merge(doy_prob, miss_days, by = "doy", all = TRUE, sort = FALSE)
doy_prob$prob[is.na(doy_prob$prob)] <- 0
doy_prob$freq[is.na(doy_prob$freq)] <- 0
doy_prob$span[is.na(doy_prob$span)] <- time_span

# discard "2012"
doy_prob$date <- substr(doy_prob$date, 6, 10)

head(doy_prob)                         
summary(doy_prob)
```

Probability of a tornado day, unsmoothed, similar to Figure 3 of Brooks et al.

```{r}
gfx_line <- ggplot(data = doy_prob, aes(x = as.numeric(doy), y = prob, group = 1))
gfx_line <- gfx_line + geom_point()
gfx_line <- gfx_line + xlab("Day of Year") + ylab("Probability")
print(gfx_line)
```

Temporal smoothing, similar to Equation 1. Make the data periodic to avoid problems with beginning and ending of data, cosistent with Brooks et al. Add bottom half the data at the top and top half of the data at the bottom. Test temporal smoothing of 5 and 15 days.

```{r}
doy_prob <- rbind(doy_prob[184:366, ], doy_prob, doy_prob[1:183, ])

sigma_t <- 5 # temporal smoother
time_wts5 <- exp(-0.5 * (c(-183:182)/ sigma_t)^2) / (sqrt(2 * pi) * sigma_t)
sigma_t <- 15 # temporal smoother
time_wts15 <- exp(-0.5 * (c(-183:182)/ sigma_t)^2) / (sqrt(2 * pi) * sigma_t)

doy_prob$freq_smooth5 <- 0
doy_prob$freq_smooth15 <- 0
for (eachDay in c(184:549)) {
  beg_index <- eachDay - 183
  end_index <- eachDay + 182
  doy_prob$freq_smooth5[eachDay] <- sum(doy_prob$freq[beg_index:end_index] * time_wts5) 
  doy_prob$freq_smooth15[eachDay] <- sum(doy_prob$freq[beg_index:end_index] * time_wts15) 
}

# discard half cycles at the top and bottom, added previously
doy_prob <- doy_prob[c(184:549), ]
# feb 29 span of 5 days does not apply after smoothing
doy_prob$prob_smooth5 <- doy_prob$freq_smooth5 / time_span
doy_prob$prob_smooth15 <- doy_prob$freq_smooth15 / time_span

summary(doy_prob)
```

Probability of a tornado day, similar to Figure 3 of Brooks et al.

```{r}
gfx_line <- ggplot(data = doy_prob, aes(x = as.numeric(doy), y = prob, group = 1))
gfx_line <- gfx_line + geom_point()
gfx_line <- gfx_line + geom_line(aes(y = prob_smooth5, colour = "red"))
gfx_line <- gfx_line + geom_line(aes(y = prob_smooth15, colour = "blue"))
gfx_line <- gfx_line + theme(legend.position = "none")
gfx_line <- gfx_line + xlab("Day of Year") + ylab("Probability")
print(gfx_line)
```

### Figure 4, Spatial Smoothing

Add temporal smoothed info to the torn 1980-99 dataset.

```{r}
doy_prob <- doy_prob[, c("date", "freq_smooth15")]
doy_prob$MONTH <- as.numeric(substr(doy_prob$date, 1, 2))
doy_prob$DAY <- as.numeric(substr(doy_prob$date, 4, 5))
doy_prob$date <- NULL
colnames(doy_prob)[1] <- "freq_t"

torn_8099 <- merge(torn_8099, doy_prob, by = c("MONTH", "DAY"), all = TRUE, sort = FALSE)
```

