Stephanie Verbout and others used data from the Storm Prediction Center to examine the evolving nature of the historical tornado record in the United States (Verbout, Stephanie M., et al. "Evolution of the US tornado database: 1954-2003." Weather and forecasting 21.1 (2006): 86-93.). 

The below code produces numbers consistent, if not identical, to those from Verbout et al. The discrepancy could be due to the differences in Tornado data used (due to regular quality control the latest data used here may not be identical to the one used by SPC).

Read raw data add column names similar to those used by [Elsner et al 2013](http://rpubs.com/jelsner/4205) and consistent with [documentation](http://www.spc.noaa.gov/gis/svrgis/). 

```{r}
library(plyr)

torn <- read.csv("data/1950-2012_torn.csv", 
                 header = FALSE, 
                 sep = ",", 
                 as.is = TRUE)

colnames(torn) <- c("OM", "YEAR", "MONTH", "DAY", "DATE", "TIME", "TIMEZONE", 
                    "STATE", "FIPS", "STATENUMBER", "FSCALE", "INJURIES", 
                    "FATALITIES", "LOSS", "CROPLOSS", "SLAT", "SLON", "ELAT", 
                    "ELON", "LENGTH", "WIDTH", "NS", "SN", "SG", "F1", "F2", 
                    "F3", "F4")
```

Tornadoes spanning multiple counties are listed separately for each county. Thus, a single tornado could appear multiple times. Identify unique tornadoes using YEAR, OM and NS. Check for uniqueness; especially, tornadoes spanning multiple years (i.e, those which begin on 12/31 and end on 1/1); need to check only those with NS > 1.

```{r}
dec31 <- subset(torn, MONTH == 12 & DAY == 31 & NS != 1)
jan01 <- subset(torn, MONTH == 1 & DAY == 1 & NS != 1)
if (nrow(dec31) > 0 & nrow(jan01) > 0) {
  stop("check! unique id assignment may not be accurate!")
}
torn$id <- paste(torn$YEAR, torn$MONTH, torn$OM, torn$NS, sep = "-")

str(torn)

head(torn)
tail(torn)
```

Function to summarize counts of unique number of tornadoes by year and month.

```{r}
count_unique_tornadoes <- function(in_df, monthly_stats = TRUE) {
  
  require(plyr)
  
  if (monthly_stats) {
    # some months dont have data; assign NAs to those months
    mon_totals <- expand.grid(MONTH = seq(1, 12), stringsAsFactors = FALSE)
    
    # number of unique tornadoes per month
    mon_torn <- ddply(.data = in_df, 
                      .variables = .(MONTH),
                      .fun = function(x_df) length(unique(x_df$id)), 
                      .drop = FALSE)
    
    mon_totals <- merge(mon_totals, mon_torn, by = "MONTH", all = TRUE)
    
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)), mon_totals$V1)
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique", month.abb)  
  } else {
    # output matrix
    out_mat <- c(nrow(in_df), length(unique(in_df$id)))
    out_mat <- matrix(out_mat, nrow = 1)
    colnames(out_mat) <- c("N_total", "N_unique")      
  }
  
  return (out_mat)
}
```

Reproduce stats on counts of tornadoes produced by Verbout et al in Figure 1 of their paper. Counts of tornadoes by F-scale.

```{r}
events_all <- ddply(.data = torn, 
                    .variables = .(YEAR), 
                    .fun = count_unique_tornadoes,
                    monthly_stats = FALSE)
  
# F1 or greater events
events_f1 <- ddply(.data = subset(torn, !(FSCALE %in% c("-9", "0"))), 
                   .variables = .(YEAR), 
                   .fun = count_unique_tornadoes,
                   monthly_stats = FALSE)
head(events_f1)

# F2 or greater events
events_f2 <- ddply(.data = subset(torn, !(FSCALE %in% c("-9", "0", "1"))), 
                   .variables = .(YEAR), 
                   .fun = count_unique_tornadoes,
                   monthly_stats = FALSE)
head(events_f2)

# F3 or greater events
events_f3 <- ddply(.data = subset(torn, !(FSCALE %in% c("-9", "0", "1", "2"))), 
                   .variables = .(YEAR), 
                   .fun = count_unique_tornadoes,
                   monthly_stats = FALSE)
head(events_f3)

# F4 or greater events
events_f4 <- ddply(.data = subset(torn, !(FSCALE %in% c("-9", "0", "1", "2", "3"))), 
                   .variables = .(YEAR), 
                   .fun = count_unique_tornadoes,
                   monthly_stats = FALSE)
head(events_f4)

event_stats <- cbind(events_all, events_f1, events_f2, events_f3, events_f4)
event_stats <- event_stats[, c(1, 3, 6, 9, 12, 15)]
colnames(event_stats) <- c("YEAR", "events_all", "F1+", "F2+", "F3+", "F4+") 

str(event_stats)

event_stats
```
